---
title: "462FinalProject"
author: "Lufei Yang"
date: "2025-04-22"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## data exploration
```{r}
library(tidyverse)
library(lubridate)
library(skimr)
library(naniar)


fl22 <- read_csv("flights2022.csv")
fl23 <- read_csv("flights2023.csv")
fl    <- bind_rows(
  fl22 %>% mutate(dataset = "2022"),
  fl23 %>% mutate(dataset = "2023")
)

dim(fl22); dim(fl23)
glimpse(fl)
skim(fl)

# Missingness 
missings <- fl %>%
  miss_var_summary() %>%
  arrange(desc(n_miss))
print(missings)

# Target distribution
fl <- fl %>%
  mutate(DEP_DEL15 = as.integer(DEP_DEL15))

fl %>%
  count(dataset, DEP_DEL15) %>%
  group_by(dataset) %>%
  mutate(pct = n / sum(n))

# Temporal patterns 
fl <- fl %>%
  mutate(
    flight_date = as.Date(FL_DATE),
    month       = factor(month(flight_date, label = TRUE), ordered = TRUE),
    weekday     = factor(wday(flight_date, label = TRUE), ordered = TRUE)
  )

# flights per month
fl %>% count(dataset, month) %>% arrange(dataset, month)

# delay rate by weekday
fl %>%
  group_by(dataset, weekday) %>%
  summarise(
    avg_delay = mean(DEP_DEL15, na.rm = TRUE),
    n_flights = n()
  ) %>%
  arrange(dataset, weekday)

# Carrier‑level summary 
fl %>%
  group_by(OP_UNIQUE_CARRIER) %>%
  summarise(
    n_flights  = n(),
    delay_rate = mean(DEP_DEL15, na.rm = TRUE)
  ) %>%
  arrange(desc(n_flights)) %>%
  slice_head(n = 10)

# Departure delay stats 
fl %>%
  filter(!is.na(DEP_DELAY)) %>%
  summarise(
    mean_delay   = mean(DEP_DELAY),
    sd_delay     = sd(DEP_DELAY),
    median_delay = median(DEP_DELAY),
    max_delay    = max(DEP_DELAY)
  )

```

The combined 2022–23 PIT departures dataset comprises roughly 137000 flights and 64 raw fields. Three post‑departure metrics (`FIRST_DEP_TIME`, `TOTAL_ADD_GTIME`, `LONGEST_ADD_GTIME`) are missing in over 99% of records—since they only exist once a flight actually departs, they leak the outcome and can be dropped. Similarly, the five “cause‑of‑delay” columns (carrier, weather, NAS, security, late‑aircraft) are about 82% missing, as they’re only populated on a delay, so they too must be excluded. By contrast, arrival details like `ARR_DELAY`, `WHEELS_OFF`, and `TAXI_IN/OUT` exceed 97% completeness and offer strong predictive signal.

Our binary target, `DEP_DEL15`, shows that 19.1% of flights in 2022 (17.4% in 2023) were delayed by at least 15minutes, with cancellations (NA) under 3.5%. Flight volume is steady at about 2000 departures per month, so seasonal effects are likely subtle; however, weekday variation is pronounced—Tuesdays and Saturdays see delay rates around 21–22%, while Mondays drop to about 17%. Carrier‑level analysis reveals a wide spread among the busiest operators: delay rates range from 11.5% (carrier YX) up to 27.1% (G4), with most majors falling between 15% and 26%. Together, these patterns suggest dropping the post‑departure and cause‑of‑delay fields, leveraging complete arrival metrics, encoding weekday effects, and grouping carriers by historical performance as key steps before model training.


```{r}
fl %>%
  group_by(weekday) %>%
  summarise(delay_rate = mean(DEP_DEL15, na.rm=TRUE)) %>%
  ggplot(aes(x=weekday, y=delay_rate)) +
    geom_col() +
    labs(title="Delay Rate by Weekday", y="P(Delay ≥15min)")

fl %>%
  group_by(DEP_TIME_BLK) %>%
  summarise(delay_rate = mean(DEP_DEL15, na.rm=TRUE), n=n()) %>%
  filter(n > 100) %>%              # drop tiny bins
  ggplot(aes(x=DEP_TIME_BLK, y=delay_rate)) +
    geom_col() +
    theme(axis.text.x = element_text(angle=45, hjust=1)) +
    labs(title="Delay Rate by Scheduled Departure Time Block", y="P(Delay ≥15min)")

fl %>%
  filter(!is.na(DEP_DELAY)) %>%
  ggplot(aes(x=factor(DISTANCE_GROUP), y=DEP_DELAY)) +
    geom_boxplot(outlier.shape=NA) +
    coord_flip() +
    labs(x="Distance Group", y="Departure Delay (min)",
         title="DEP_DELAY Distribution by Distance Group")

fl %>%
  filter(!is.na(TAXI_OUT), !is.na(DEP_DELAY)) %>%
  ggplot(aes(x=TAXI_OUT, y=DEP_DELAY)) +
    geom_point(alpha=0.2) +
    geom_smooth(method="lm", se=FALSE) +
    labs(title="DEP_DELAY vs. Taxi-Out Time", x="Taxi-Out (min)", y="Departure Delay (min)")

```
From plots:

1. Weekday effects
Tuesdays still top the chart (~21% chance of ≥15min delay), with Wednesdays lowest (~16%). There’s a clear, roughly 5pp swing across the week—definitely worth encoding as a factor.

2. Time‑of‑day ramp
Delay probability climbs steadily from the early morning (only ~6% between midnight–6AM) to the evening peak (~33% for 8–9PM departures). A simple 24‑level DEP_TIME_BLK factor (or even a numeric “hour of day”) should capture a large chunk of our signal.

3. Distance group vs. delay
Each successive distance bucket shows a slight upward shift in median departure delay, but the variance is so huge (and the outliers so extreme) that I’d treat distance as a continuous feature—perhaps log‑transforming the raw miles rather than using the coarse groups.

4. Taxi‑out vs. departure delay
Beyond the obvious outliers (flights that sit on the tarmac forever), there’s almost no linear relationship. Taxi‑out time may still carry some information (e.g. airport congestion), but it’s not a strong standalone predictor and may be better used in interaction (e.g. high‐traffic evening slots).

## baseline
```{r}
library(tidyverse)
library(lubridate)
library(pROC)

fl22 <- read_csv("flights2022.csv")
fl23 <- read_csv("flights2023.csv")

fl <- bind_rows(
  fl22 %>% mutate(dataset = "2022"),
  fl23 %>% mutate(dataset = "2023")
) %>%
  mutate(
    DEP_DEL15   = as.integer(DEP_DEL15),
    flight_date = as.Date(FL_DATE, format = "%m/%d/%Y")
  )

feats <- fl %>%
  filter(!is.na(DEP_DEL15)) %>%
  transmute(
    DEP_DEL15,
    TAIL_NUM,
    flight_date = as.Date(FL_DATE, format = "%m/%d/%Y"),
    wd          = factor(wday(flight_date, label = TRUE), ordered = TRUE),
    hr          = CRS_DEP_TIME %/% 100,
    log_dist    = log(DISTANCE),
    taxi_out    = TAXI_OUT,
    carrier     = OP_UNIQUE_CARRIER
  )

set.seed(42)
train_ix <- sample(nrow(feats), size = 0.7 * nrow(feats))
train    <- feats[train_ix, ]
valid    <- feats[-train_ix, ]

enc <- train %>%
  group_by(carrier) %>%
  summarise(carrier_rate = mean(DEP_DEL15), .groups = "drop")

base_rate <- mean(train$DEP_DEL15)

train <- train %>%
  left_join(enc, by = "carrier")

valid <- valid %>%
  left_join(enc, by = "carrier") %>%
  mutate(carrier_rate = replace_na(carrier_rate, base_rate))

mod_glm <- glm(
  DEP_DEL15 ~ wd + hr + log_dist + taxi_out + carrier_rate,
  data   = train,
  family = binomial
)

pred_p  <- predict(mod_glm, newdata = valid, type = "response")
roc_obj <- roc(valid$DEP_DEL15, pred_p)
auc_val <- auc(roc_obj)
cat("Logistic AUC:", round(auc_val, 4), "\n")

```
```{r}
library(tidyverse)
library(lubridate)
library(xgboost)
library(pROC)

# prepare previous‐flight arrival delays
last_arr <- fl %>%
  filter(!is.na(ARR_DELAY)) %>%
  transmute(
    TAIL_NUM,
    flight_date    = as.Date(FL_DATE, format = "%m/%d/%Y"),
    hr              = CRS_DEP_TIME %/% 100,
    prev_arr_delay = ARR_DELAY
  ) %>%
  arrange(TAIL_NUM, flight_date, hr) %>%
  group_by(TAIL_NUM, flight_date, hr) %>%
  slice_tail(n = 1) %>%
  ungroup()

# merge into train/valid and impute
train2 <- train %>%
  left_join(last_arr, by = c("TAIL_NUM","flight_date","hr")) %>%
  mutate(
    prev_arr_delay = replace_na(prev_arr_delay, 0),
    taxi_out       = replace_na(taxi_out, median(taxi_out, na.rm = TRUE))
  )

valid2 <- valid %>%
  left_join(last_arr, by = c("TAIL_NUM","flight_date","hr")) %>%
  mutate(
    prev_arr_delay = replace_na(prev_arr_delay, 0),
    taxi_out       = replace_na(taxi_out, median(train2$taxi_out, na.rm = TRUE))
  )

# build model matrices
x_train <- model.matrix(~ wd + hr + log_dist + taxi_out + carrier_rate + prev_arr_delay - 1, data = train2)
x_valid <- model.matrix(~ wd + hr + log_dist + taxi_out + carrier_rate + prev_arr_delay - 1, data = valid2)

dtrain <- xgb.DMatrix(data = x_train, label = train2$DEP_DEL15)
dvalid <- xgb.DMatrix(data = x_valid, label = valid2$DEP_DEL15)

params <- list(
  objective   = "binary:logistic",
  eval_metric = "auc",
  eta         = 0.1,
  max_depth   = 4
)

bst <- xgb.train(
  params,
  dtrain,
  nrounds               = 100,
  watchlist             = list(train = dtrain, valid = dvalid),
  early_stopping_rounds = 10,
  verbose               = 0
)

params <- list(
  objective        = "binary:logistic",
  eval_metric      = "auc",
  eta              = 0.05,      # halve the learning rate
  max_depth        = 4,         # keep trees shallow
  subsample        = 0.8,       # sample 80% of rows per tree
  colsample_bytree = 0.7,       # sample 70% of columns per tree
  lambda           = 1.0,       # L2 regularization
  alpha            = 0.5        # L1 regularization
)

bst <- xgb.train(
  params                = params,
  data                  = dtrain,
  nrounds               = 200,                # allow more rounds, but ES will stop early
  watchlist             = list(train=dtrain, valid=dvalid),
  early_stopping_rounds = 10,
  record                = TRUE,
  verbose               = 1
)

# replot the learning curve to confirm the gap narrowed
log <- as.data.frame(bst$evaluation_log)
library(ggplot2)
ggplot(log, aes(x=iter)) +
  geom_line(aes(y=train_auc, color="Train")) +
  geom_line(aes(y=valid_auc, color="Valid")) +
  labs(title="Regularized XGB Learning Curve",
       x="Boost Round", y="AUC") +
  scale_color_manual("",values=c("Train"="steelblue","Valid"="firebrick")) +
  geom_vline(xintercept=bst$best_iteration, linetype=2) +
  theme_minimal()


pred_xgb <- predict(bst, dvalid)
auc_xgb  <- auc(roc(valid2$DEP_DEL15, pred_xgb))
cat("XGBoost AUC:", round(auc_xgb, 4), "\n")

```


## Analysis on xgboost
```{r}
library(ggplot2)
library(pROC)
library(xgboost)

# ROC curve comparison 
# compute ROC objects
roc_glm <- roc(valid$DEP_DEL15, pred_p)
roc_xgb <- roc(valid2$DEP_DEL15, pred_xgb)  

# build a data frame for ggplot
roc_df <- bind_rows(
  tibble(
    fpr = 1 - roc_glm$specificities,
    tpr = roc_glm$sensitivities,
    model = "Logistic"
  ),
  tibble(
    fpr = 1 - roc_xgb$specificities,
    tpr = roc_xgb$sensitivities,
    model = "XGBoost"
  )
)

ggplot(roc_df, aes(x = fpr, y = tpr, color = model)) +
  geom_line(size = 1) +
  geom_abline(lty = 2) +
  labs(
    title = "ROC Curves: Logistic vs XGBoost",
    x = "False Positive Rate",
    y = "True Positive Rate"
  )

# Feature importance from XGBoost
imp <- xgb.importance(model = bst)
# plot top 10
top_imp <- imp[1:10, ]
ggplot(top_imp, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col() +
  coord_flip() +
  labs(
    title = "Top 10 XGBoost Features by Gain",
    x = NULL, y = "Relative Importance (Gain)"
  )

# Calibration plot for XGBoost 
valid2$pred_bin <- cut(pred_xgb, breaks = seq(0,1, by = .1), include.lowest = TRUE)
calib <- valid2 %>%
  group_by(pred_bin) %>%
  summarise(
    mean_pred = mean(pred_xgb),
    obs_rate  = mean(DEP_DEL15),
    n         = n()
  )

ggplot(calib, aes(x = mean_pred, y = obs_rate)) +
  geom_point(aes(size = n), alpha = 0.6) +
  geom_abline(lty = 2) +
  labs(
    title = "Calibration Plot: XGBoost",
    x = "Mean Predicted Probability",
    y = "Observed Delay Rate"
  ) +
  scale_size_area("Bin size")

# 4. Partial dependence for previous‐arrival delay
library(pdp)
pd_prev <- partial(bst, pred.var = "prev_arr_delay", train = x_train, grid.resolution = 20)
autoplot(pd_prev) +
  labs(
    title = "Partial Dependence: Previous Flight Arrival Delay",
    x = "Prev Arr Delay (min)",
    y = "Predicted Delay Probability"
  )

```

```{r}
# =============================
# Final Test‐Set Pipeline & Submission
# =============================

library(tidyverse)
library(lubridate)
library(xgboost)

# — assume these objects are already in your workspace from training:
#    bst         : your fitted xgboost model
#    enc         : data frame (carrier, carrier_rate)
#    base_rate   : overall mean(train$DEP_DEL15)
#    last_arr    : data frame (TAIL_NUM, flight_date, hr, prev_arr_delay)
#    train       : your training data.frame (for wd levels & taxi_out median)
#    valid2      : your validation data.frame (for test.acc & pred_xgb)

# 1. Read the guess file, forcing key columns to numeric
test_guess <- read_csv(
  "flights2024_guess.csv",
  col_types = cols(
    TAIL_NUM          = col_character(),
    FL_DATE           = col_character(),
    OP_UNIQUE_CARRIER = col_character(),
    CRS_DEP_TIME      = col_double(),
    DISTANCE          = col_double(),
    TAXI_OUT          = col_double(),
    .default          = col_guess()
  )
)

# 2. Feature‐engineer the test set
test_feat <- test_guess %>%
  mutate(
    flight_date    = as.Date(FL_DATE, "%m/%d/%Y"),
    wd             = factor(
                       wday(flight_date, label = TRUE),
                       levels = levels(train$wd),
                       ordered = TRUE
                     ),
    hr             = CRS_DEP_TIME %/% 100,
    log_dist       = log(DISTANCE),
    taxi_out       = TAXI_OUT,
    carrier        = OP_UNIQUE_CARRIER
  ) %>%
  # carrier mean‐encoding
  left_join(enc, by = "carrier") %>%
  mutate(
    carrier_rate = coalesce(carrier_rate, base_rate)
  ) %>%
  # previous‐flight arrival delay
  left_join(
    last_arr,
    by = c("TAIL_NUM", "flight_date", "hr")
  ) %>%
  mutate(
    prev_arr_delay = replace_na(prev_arr_delay, 0),
    taxi_out       = replace_na(taxi_out, median(train$taxi_out, na.rm = TRUE))
  )

# 3. Generate predictions
x_test <- model.matrix(
  ~ wd + hr + log_dist + taxi_out + carrier_rate + prev_arr_delay - 1,
  data = test_feat
)
delay.guesses <- predict(bst, xgb.DMatrix(x_test))

# 4. Estimate test‐set accuracy (using your validation proxy)
test.acc <- mean((pred_xgb >= 0.5) != valid2$DEP_DEL15)

# 5. Define your team name and save
team.name <- "Team462"

save(
  list = c("delay.guesses", "test.acc", "team.name"),
  file = "Team462final.RData"
)


```

```{r}
load("Team462final.RData")
length(delay.guesses)   # should now be 8723 (or however many test rows)
head(delay.guesses)
summary(delay.guesses)
View(delay.guesses)

print(team.name)
print(test.acc)
nrow(read_csv("flights2024_guess.csv"))
length(delay.guesses)
tail(delay.guesses)

```

